Novel improvements in the Material Point Method to face dynamic problems: Local-maximum entropy approximation and an explicit predictor-corrector scheme. \thanks{Funding: This study was funded by Agustín de Betancourt Foundation (grant number 262390106114).}

### 2020-03-10 main.tex
Cambios 10-3-2020

% Divergence
\newcommand{\Div}[1]{
  \ensuremath{\nabla \cdot {#1}}
}
% Gradient
\newcommand\Grad[1]{\nabla{#1}}
\newcommand\GradS[1]{\nabla^s{#1}}
\newcommand\GradT[1]{\nabla^T{#1}}

@@

Three kind of subscript or superscript are used within paper. The
subscript $\qed_p$ is used to define a particle variable. While the
subscript $\qed_I$ is reserved in this notation for denoting nodal
variables. And finally, the superscript $\qed^{\psi}$ involves a virtual
magnitude. For the operators, the convention is : $\dot{\qed}$ and
$\ddot{\qed}$ for the first and second time derivative, $\qued \otimes
\qued$ means the dyadic operator, $(\cdot)$ and $(\colon)$ means the first
and second contraction of a tensor, $\Div{\qed}$
denotes the divergence operator, and finally $\Grad{\qed}$ and
$\GradS{\qed}$ denotes the gradient and its symmetric part.

@@

\subsection{Governing equations }
\label{sec:governing-equations}
In the MPM approach a continuum is considered. Let define a
$\varOmega$ domain occupied by an elastic body, and $\partial
\varOmega$ the boundary of the domain, where $\partial
\varOmega = \Gamma_d \bigcup \Gamma_n$ and $\Gamma_d \cap  \Gamma_n =
\emptyset$. Note that $\varGamma_d$ is reserved for Dirichlet boundary
conditions, and  $\varGamma_n$ for the Neumann. The behaviour of the continuum can be described by a set of governing equations which are
(i) the balance of momentum equation
\begin{equation}
  \label{eq:Balance-momentum}
\rho \vect{a} = \Div{\tens{\sigma}} + \rho \vect{b},
\end{equation}
(ii) the compatibility equation with ensure that the velocity
field is compatible with the Cauchy strain field,
\begin{equation}
  \label{eq:Compatibility-equation}
  \dot{\tens{\varepsilon}} = \GradS{\vect{v}} = \frac{1}{2}\left(
    \Grad{\vect{v}} +  \GradT{\vect{v}} \right),
\end{equation}
(iii) the constitutive equation, which relates the rate of Cauchy strain
tensor with the rate of Cauchy stress tensor,
\begin{equation}
  \label{eq:Constitutive-equation}
\dot{\tens{\sigma}} = \tens{D} \colon \dot{\tens{\varepsilon}},
\end{equation}
and finally (iv) the mass conservation, which can be obtained by
setting to zero the total derivative of the density field,
\begin{equation}
  \label{eq:Rho-material-derivative}
  \frac{D \rho}{D t} = \dot{\rho} + \rho \Div\vect{v}} = 0.
\end{equation}

\subsection{Variational formulation}
\label{sec:variational-formulation}
To write the variational statement of the problem, let us define a
virtual velocity field such that
\begin{equation}
  \label{eq:Hilbert-space}
  \vect{v}^{\psi} \in \mathcal{H}^1_0(\Omega) = \{ \vect{v}^{\psi} \in
  \mathcal{H}^1 \mid \vect{v}^{\psi} = \vect{0}\ \text{on}\ \Gamma_d \}.
\end{equation}
And which satisfies that the Cauchy sequences are convergent in $\varOmega$
\begin{equation}
  \label{eq:cauchy-secuence}
  \Integral{3}{\vect{v}^{\psi}} < \infty\ \quad\text{and}\quad \Integral{3}{\tens{\varepsilon}^{\psi}} < \infty
\end{equation}
The principle of virtual work states that the equilibrium solution to
the boundary value problem of elasticity is the function $\vect{v} \in
\mathcal{H}^1_0$ such that, for $\vect{v}^{\psi} \in
\mathcal{H}^1_0$, the following holds:
\begin{equation}
  \label{eq:BalanceMomentum_wf}
  \Integral{3}{\rho\ \left( \Deriv{\vec{v}}{t}\ - \vec{b} \right) \cdot \vec{v}^{\psi}} =
  \Integral{2}{\vec{t}\ \cdot \vec{v}^{\psi}} - \Integral{3}{\tens{\sigma} \cdot
    \tens{\varepsilon}^{\psi}},
\end{equation}\\
therefore (\ref{eq:BalanceMomentum_wf}) is the balance of momentum formulated in
its weak form.

@@

Each material point is assigned with initial values of position,
velocity, mass, volume and stress denoted by $\vec{x}_p$,
$\vec{v}_p$, $m_p$,  $V_p$ and $\tens{\sigma}_p$, but also the
virtual field $\vect{v}^{\psi}_{p}$. So employing the definition of the
material integral, where we recover the Riemann integral definition as
an addition  of a finite set of points. Therefore individual terms in
\eqref{eq:BalanceMomentum_wf} are solved as follows. 
\begin{itemize}
\item Acceleration forces :
\begin{equation}
    \label{eq:particle_acceleration_forces}
    \Integral{3}{\rho\ \vec{a}\ \vect{v}^{\psi}} =& \sum^{N_p}_{p}
   \dot{\vec{v}}_{p}\ \vect{v}^{\psi}_{p}\ m_p.
  \end{equation}\\
\item Internal forces :
  \begin{equation}
    \label{eq:particle_internal_forces}
    \Integral{3}{\tens{\sigma}\ \cdot \Grad{\vec{\psi}}} = \sum^{N_p}_{p}
   \tens{\sigma}_{p}\ \cdot \Grad{\tens{\psi}_{p}}\ V_p.
  \end{equation}\\
\item Body forces :
\begin{equation}
  \label{eq:particle_body_forces}
  \Integral{3}{\rho\ \vec{b}\ \vect{v}^{\psi} } = \sum^{N_p}_{p}
  \vec{b}_{p}\ \vect{v}^{\psi}_p\ m_p.
\end{equation}\\
\item Loads :
\begin{equation}
  \begin{aligned}
    \label{eq:particle_load_forces}
    \Integral{2}{\vec{t}\ \vect{v}^{\psi}} = \Integral{2}{\rho\
      \vec{t}^s\ \vect{v}^{\psi}} = \sum^{N_p}_{p} \vec{t}^s_{p}\ \vect{v}^{\psi}_{p}\ h^{-1}\ m_p ,
  \end{aligned} 
\end{equation}
\end{itemize}
where $h$ is the thickness of the continuum in a 2D case. Here is
where the second discretization procedure appears. A background mesh
composed by a finite set of grid points with coordinates $\vec{x}_I\
, I = 1,2\ldots ,N_n$, is generated. Where $N_n$ is the number of grid
nodes. This mesh is employed as a support to compute gradients and divergences.

Introducing \eqref{eq:particle_acceleration_forces},
\eqref{eq:particle_internal_forces}, \eqref{eq:particle_body_forces}, 
and \eqref{eq:particle_load_forces} in \eqref{eq:BalanceMomentum_wf},
and approximating the real and virtual velocity  field of the particle
$p$ as
\begin{equation}
  \label{eq:Approx-velocity-mesh}
  \vec{v}_{p} = \sum^{Nn}_I N_I(\vec{x}_p) \vect{v}_I
  \quad \text{and} \quad \vec{v}^{\psi}_{p} = \sum^{Nn}_I N_I(\vec{x}_p) \vect{1},
\end{equation}
we reach to the particle balance of forces of the continuum,
\begin{equation}
  \label{eq:particle_balance_forces3}
  \dot{\vec{p}}_{I}= \tens{m}_{IJ}\dot{\vec{v}}_{J} = \vec{f}_{I}^{int} + \vec{f}_{I}^{ext},
\end{equation}
where $\dot{\vec{p}}_{I}$ is the rate of momentum at grid node $I$, the nodal mass matrix $\tens{m}_{IJ}$ is,
\begin{equation}
  \label{eq:particle_nod_mass}
  \tens{m}_{IJ} =
  \sum^{Np}_{p} N_{Ip} m_p N_{Jp}.
\end{equation}
To improve the computational efficiency and stability, the nodal mass matrix
\eqref{eq:particle_nod_mass} can be substituted by the lumped mass
matrix $\tens{m}_{IJ}^{lumped}$.
Later, internal and external forces are computed as follows,
\begin{equation}
  \label{eq:nodal_internal_forces}
  \vec{f}_{I}^{int} = - \sum^{N_p}_{p} \tens{\sigma}_{p} \cdot \Grad{N_{Ip}} \frac{m_p}{\rho_p}
\end{equation}
\begin{equation}
  \label{eq:nodal_external_forces}
  \vec{f}_{I}^{ext} = \sum^{N_p}_{p} N_{Ip} \vec{b}_{p} m_p  + \sum^{N_p}_{p}
  N_{Ip} \vec{t}^s_{p} m_p h^{-1} 
\end{equation}
where $\tens{\sigma}_{p} = \tens{\sigma}_{p}(\tens{\varepsilon}_{p})$
is the particle $p$ stress field, which can be integrated employing
the suitable constitutive model. The strain tensor is updated employing the rate of stress tensor $\dot{ \tens{\varepsilon}}_{p}$ used to update the
strain tensor is as follows \eqref{eq:IncrStrainPoint}.
\begin{equation}
  \label{eq:IncrStrainPoint}
  \dot{\tens{\varepsilon}_{p}} = \frac{\Delta
    \tens{\varepsilon}_{p}}{\Delta t} =
  \frac{1}{2} \sum^{Nn}_I \left[\Grad{N_{Ip}}\ \otimes \vec{v}_{I} + \vec{v}_{I} \otimes
    \Grad{N_{Ip}}\ \right].
\end{equation}
Next, imposing $\frac{D \rho}{D t} = 0$, we ensures the mass
conservation and update the density field.
\begin{equation}
  \label{eq:MassConservation}
\dot{\rho} = - \rho\ \mathit{tra} \left( \dot{\tens{\varepsilon}} \right)
\end{equation}

>>>>

% Divergence
\newcommand{\Div}[1]{
  \ensuremath{div({#1})}
}
% Gradient
\newcommand\Grad[1]{grad({#1})}
\newcommand\GradS[1]{grad^s({#1})}
\newcommand\GradT[1]{grad^T({#1})}

@@

Three kind of subscript or superscript are used within paper. The
subscript $\qed_p$ is used to define a particle variable. While the
subscript $\qed_I$ is reserved in this notation for denoting nodal
variables. And finally, the superscript $\qed^{\psi}$ involves a virtual
magnitude. For the operators, the convention is : $\dot{\qed}$ and
$\ddot{\qed}$ for the first and second time derivative, $\qued \otimes
\qued$ means the dyadic operator, $(\cdot)$ and $(\colon)$ means the first
and second contraction of a tensor, $\Div{\qed}$
denotes the divergence operator, and finally $\Grad{\qed}$ and
$\GradS{\qed}$ denotes the gradient and its symmetric part. Einstain
subscripts convention is adopted so repeated index means addition.

@@

\subsection{Governing equations }
\label{sec:governing-equations}
In the MPM approach a continuum is considered. Let define a
$\varOmega$ domain occupied by an elastic body, and $\partial
\varOmega$ the boundary of the domain, where $\partial
\varOmega = \Gamma_d \bigcup \Gamma_n$ and $\Gamma_d \cap  \Gamma_n =
\emptyset$. Note that $\varGamma_d$ is reserved for Dirichlet boundary
conditions, and  $\varGamma_n$ for the Neumann. In a inifinitesimal
strain context, the behaviour of the continuum can be described by the
following set of equations:
(i) the balance of momentum equation
\begin{equation}
  \label{eq:Balance-momentum}
\rho \vect{a} = \Div{\tens{\sigma}} + \rho \vect{b},
\end{equation}
(ii) the compatibility equation with ensure that the velocity
field is compatible with the Cauchy strain field,
\begin{equation}
  \label{eq:Compatibility-equation}
  \dot{\tens{\varepsilon}} = \GradS{\vect{v}},
\end{equation}
(iii) the constitutive equation, which relates the rate of Cauchy strain
tensor with the rate of Cauchy stress tensor,
\begin{equation}
  \label{eq:Constitutive-equation}
\dot{\tens{\sigma}} = \tens{D} \colon \dot{\tens{\varepsilon}},
\end{equation}
and finally (iv) the mass conservation, which can be obtained by
setting to zero the total derivative of the density field,
\begin{equation}
  \label{eq:Rho-material-derivative}
  \frac{D \rho}{D t} = \dot{\rho} + \rho \Div{\vect{v}} = 0.
\end{equation}

\subsection{Variational formulation}
\label{sec:variational-formulation}
To write the variational statement of the problem, let us define a
virtual velocity field such that
\begin{equation}
  \label{eq:Hilbert-space}
  \vect{v}^{\psi} \in \mathcal{H}^1_0(\Omega) = \{ \vect{v}^{\psi} \in
  \mathcal{H}^1 \mid \vect{v}^{\psi} = \vect{0}\ \text{on}\ \Gamma_d \}.
\end{equation}
And which satisfies that the Cauchy sequences are convergent in $\varOmega$
\begin{equation}
  \label{eq:cauchy-secuence}
  \Integral{3}{\vect{v}^{\psi}} < \infty\ \quad\text{and}\quad \Integral{3}{\tens{\varepsilon}^{\psi}} < \infty
\end{equation}
The principle of virtual work states that the equilibrium solution to
the boundary value problem of elasticity is the function $\vect{v} \in
\mathcal{H}^1_0$ such that, for $\vect{v}^{\psi} \in
\mathcal{H}^1_0$, the following holds:
\begin{equation}
  \label{eq:BalanceMomentum_wf}
  \Integral{3}{\rho\ \left( \Deriv{\vec{v}}{t}\ - \vec{b} \right) \cdot \vec{v}^{\psi}} =
  \Integral{2}{\vec{t}\ \cdot \vec{v}^{\psi}} - \Integral{3}{\tens{\sigma} \colon
    \tens{\varepsilon}^{\psi}},
\end{equation}\\
therefore (\ref{eq:BalanceMomentum_wf}) is the balance of momentum formulated in
its weak form.

@@

Each material point is assigned with initial values of position,
velocity, mass, volume and stress denoted by $\vec{x}_p$,
$\vec{v}_p$, $m_p$,  $V_p$ and $\tens{\sigma}_p$, but also the
virtual field $\vect{v}^{\psi}_{p}$. So employing the definition of the
material integral, where we recover the Riemann integral definition as
an addition  of a finite set of points. Therefore individual terms in
\eqref{eq:BalanceMomentum_wf} are solved as follows. 
\begin{itemize}
\item Acceleration forces :
\begin{equation}
    \label{eq:particle_acceleration_forces}
    \Integral{3}{\rho\ \vec{a} \cdot \vect{v}^{\psi}} =
    \dot{\vec{v}}_{p} \cdot \vect{v}^{\psi}_{p}\ m_p.
  \end{equation}\\
\item Internal forces :
  \begin{equation}
    \label{eq:particle_internal_forces}
    \Integral{3}{\tens{\sigma}\ \colon \tens{\varepsilon}^{\psi}} =
   \tens{\sigma}_{p}\ \colon \tens{\varepsilon}^{\psi}_p \ V_p.
  \end{equation}\\
\item Body forces :
\begin{equation}
  \label{eq:particle_body_forces}
  \Integral{3}{\rho\ \vec{b} \cdot \vect{v}^{\psi} } = 
  \vec{b}_{p} \cdot \vect{v}^{\psi}_p\ m_p.
\end{equation}\\
\item Loads :
\begin{equation}
  \begin{aligned}
    \label{eq:particle_load_forces}
    \Integral{2}{\vec{t}\ \vect{v}^{\psi}} = \Integral{2}{\rho\
      \vec{t}^s \cdot \vect{v}^{\psi}} = \vec{t}^s_{p} \cdot \vect{v}^{\psi}_{p}\ h^{-1}\ m_p ,
  \end{aligned} 
\end{equation}
\end{itemize}
where $h$ is the thickness of the continuum in a 2D case. Here is
where the second discretization procedure appears. A background mesh
composed by a finite set of grid points with coordinates $\vec{x}_I\
, I = 1,2\ldots ,N_n$, is generated. Where $N_n$ is the number of grid
nodes. This mesh is employed as a support to compute gradients and divergences.

Introducing \eqref{eq:particle_acceleration_forces},
\eqref{eq:particle_internal_forces}, \eqref{eq:particle_body_forces}, 
and \eqref{eq:particle_load_forces} in \eqref{eq:BalanceMomentum_wf},
and approximating the real and virtual velocity  field of the particle
$p$ as
\begin{equation}
  \label{eq:Approx-velocity-mesh}
  \vec{v}_{p} = N_I(\vec{x}_p) \vect{v}_I
  \quad \text{and} \quad \vec{v}^{\psi}_{p} = N_I(\vec{x}_p) \vect{v}_I^{\psi},
\end{equation}
\begin{equation}
  \label{eq:Approx-gradvelocity-mesh}
  \Grad{\vect{v}}_p = \vect{v}_I N_I(\vec{x}_p)
  \quad \text{and} \quad \vec{v}^{\psi}_{p} = \sum^{Nn}_I N_I(\vec{x}_p) \vect{v}_I^{\psi},
\end{equation}
we reach to the particle balance of forces of the continuum,
\begin{equation}
  \label{eq:particle_balance_forces3}
  \dot{\vec{p}}_{I}= \tens{m}_{IJ}\dot{\vec{v}}_{J} = \vec{f}_{I}^{int} + \vec{f}_{I}^{ext},
\end{equation}
where $\dot{\vec{p}}_{I}$ is the rate of momentum at grid node $I$, the nodal mass matrix $\tens{m}_{IJ}$ is,
\begin{equation}
  \label{eq:particle_nod_mass}
  \tens{m}_{IJ} =
  \sum^{Np}_{p} N_{Ip} m_p N_{Jp}.
\end{equation}
To improve the computational efficiency and stability, the nodal mass matrix
\eqref{eq:particle_nod_mass} can be substituted by the lumped mass
matrix $\tens{m}_{IJ}^{lumped}$.
Later, internal and external forces are computed as follows,
\begin{equation}
  \label{eq:nodal_internal_forces}
  \vec{f}_{I}^{int} = - \sum^{N_p}_{p} \tens{\sigma}_{p} \cdot \Grad{N_{Ip}} \frac{m_p}{\rho_p}
\end{equation}
\begin{equation}
  \label{eq:nodal_external_forces}
  \vec{f}_{I}^{ext} = \sum^{N_p}_{p} N_{Ip} \vec{b}_{p} m_p  + \sum^{N_p}_{p}
  N_{Ip} \vec{t}^s_{p} m_p h^{-1} 
\end{equation}
where $\tens{\sigma}_{p} = \tens{\sigma}_{p}(\tens{\varepsilon}_{p})$
is the particle $p$ stress field, which can be integrated employing
the suitable constitutive model. The strain tensor is updated employing the rate of stress tensor $\dot{ \tens{\varepsilon}}_{p}$ used to update the
strain tensor is as follows \eqref{eq:IncrStrainPoint}.
\begin{equation}
  \label{eq:IncrStrainPoint}
  \dot{\tens{\varepsilon}_{p}} = \frac{\Delta
    \tens{\varepsilon}_{p}}{\Delta t} =
  \frac{1}{2} \sum^{Nn}_I \left[\Grad{N_{Ip}}\ \otimes \vec{v}_{I} + \vec{v}_{I} \otimes
    \Grad{N_{Ip}}\ \right].
\end{equation}
Next, imposing $\frac{D \rho}{D t} = 0$, we ensures the mass
conservation and update the density field.
\begin{equation}
  \label{eq:MassConservation}
\dot{\rho} = - \rho\ \mathit{tra} \left( \dot{\tens{\varepsilon}} \right)
\end{equation}
